Elite orientation of an education system and economic growth

### 2024-04-26 paper/main.tex
should be done i think

Separate regression for each year. Does not make the assumptions above. We modify the equation above to remove fixed effects.
\begin{equation}
    Y_{i, t} = \beta_0 + \lambda E_{i, t} + \alpha I_{i, t} + \epsilon_{i, t}
\end{equation}
\subsection{Interpretion and goals}
This paper does not aim to make any causal inference as it is complicated by omitted variable bias and the reality the elite academic orientation variables are merely instruments for unobservable characteristics. GDP per capita growth obviously has many contributing factors, and so a simple scatter plot (see figure \ref{fig:relationships}) is insufficient at determining whether a statistical relationship exists.
\begin{figure}[H]
    \caption{Relationships of variables with GDP per capita growth}
    \centering
    \includegraphics[width=0.6\textwidth]{../charts/relationships.png}
    \label{fig:relationships}
\end{figure}
By running specified regression models, we can examine the relationship while considering controls. It is clear from the picture above that we must consider this \textit{conditional} relationship. The regression coefficients on the elite orientation variables will indicate the size, direction, and significance of elite orientation condition on the controls included in the model. Comparing different model specifications will also let us examine the different assumptions made by each (e.g., time heterogeneity, within/between variation). The next section will discuss the interpretations and results with this in mind.
\section{Results and Interpretation}
\subsection{Panel Regressions}
We begin by estimating the panel regression specifications given above. The results are shown below in Table \ref{table:panel}.
    \caption{Panel regression results}
    \label{table:panel}
        \begin{tabular}{@{\extracolsep{5pt}}lcccc}
            PISA Math in global P99 & -31.228$^{*}$ & -38.928$^{*}$ & -34.569$^{*}$ & -115.492$^{***}$ \\
            & (16.314) & (20.802) & (17.850) & (36.599) \\
            IMO score per log population & 10.674$^{*}$ & 4.924$^{}$ & 3.122$^{}$ & -11.158$^{}$ \\
            & (6.363) & (7.470) & (6.441) & (15.728) \\
            ARWU insitutions & -166.641$^{**}$ & -162.705$^{*}$ & -176.049$^{**}$ & -159.960$^{}$ \\
            & (68.800) & (86.871) & (72.281) & (112.884) \\
            GDP per capita & & -0.001$^{}$ & -0.001$^{}$ & 0.003$^{}$ \\
            & & (0.001) & (0.001) & (0.003) \\
            PISA Math & & 0.619$^{}$ & 0.425$^{}$ & 0.178$^{}$ \\
            & & (0.716) & (0.684) & (1.735) \\
            Primary School Completion Rate & & -12.285$^{***}$ & -6.467$^{*}$ & -3.713$^{}$ \\
            & & (4.104) & (3.586) & (4.965) \\
            Lower Sec. Completion Rate & & -0.425$^{}$ & -0.953$^{}$ & -0.194$^{}$ \\
            & & (3.533) & (2.958) & (4.475) \\
            Upper Sec. Completion Rate & & 6.987$^{**}$ & 5.721$^{**}$ & 3.474$^{}$ \\
            & & (2.843) & (2.375) & (3.707) \\
            Democracy Rating & & 3.880$^{}$ & -5.350$^{}$ & 53.356$^{}$ \\
            & & (17.079) & (14.344) & (62.249) \\
            Population & & -0.000$^{}$ & -0.000$^{}$ & 0.000$^{}$ \\
            & & (0.000) & (0.000) & (0.000) \\
            Time Effects & No & No & Yes & Yes \\
            Fixed Effects & No & No & No & Yes \\
            Controls & No & Yes & Yes & Yes \\
            Entities & 89 & 89 & 89 & 89 \\
            Observations & 440 & 440 & 440 & 440 \\
            $R^2$ & 0.037 & 0.077 & 0.374 & 0.553 \\
            Adjusted $R^2$ & 0.031 & 0.056 & 0.351 & 0.414 \\
            Residual Std. Error & 445.700 (df=436) & 439.888 (df=429) & 364.790 (df=423) & 346.511 (df=335) \\
            F Statistic & 5.616$^{***}$ (df=3; 436) & 3.589$^{***}$ (df=10; 429) & 15.813$^{***}$ (df=16; 423) & 3.983$^{***}$ (df=104; 335) \\
            \textit{Note:} & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\

@@

\subsection{Does IMO performance measure something else?}
To explore whether IMO performance indeed captures a different set of unobservable factors (the cultural and institutional/structural elite orientation of an education system) compared to PISA math 99 share and ARWU institution count. Principle Components Analysis (PCA) may help analyze this by projecting the 3 variables onto 2 orthogonal vectors.

@@

The relationships between the controls and the elite orientation variables are likely to have changed as the XGBoost model has to some extent forced the existence of a relationship.

>>>>

\subsection{Yearly Regressions}
Separate regression for each year. Does not make the assumptions above. We modify the equation above to remove fixed effects.
\begin{equation}
    Y_{i, t} = \beta_0 + \lambda E_{i, t} + \alpha I_{i, t} + \epsilon_{i, t}
\end{equation}

\subsection{Without PISA variables}
A larger sample of countries (with a similar specification as the panel regression above) is considered by removing the PISA variables which limit the rows available.
This may help reduce standard errors and illustrate whether:
\begin{itemize}
    \item The results hold on a larger set of countries and years (minus the PISA variables)
    \item Whether lower standard errors can be achieved
\end{itemize}

\subsection{Interpretion and goals}
This paper does not aim to make any causal inference as it is complicated by omitted variable bias and the reality the elite academic orientation variables are merely instruments for unobservable characteristics. GDP per capita growth obviously has many contributing factors, and so a simple scatter plot (see figure \ref{fig:relationships}) is insufficient at determining whether a statistical relationship exists.
\begin{figure}[H]
    \caption{Relationships of variables with GDP per capita growth}
    \centering
    \includegraphics[width=0.6\textwidth]{../charts/relationships.png}
    \label{fig:relationships}
\end{figure}
By running specified regression models, we can examine the relationship while considering controls. It is clear from the picture above that we must consider this \textit{conditional} relationship. The regression coefficients on the elite orientation variables will indicate the size, direction, and significance of elite orientation condition on the controls included in the model. Comparing different model specifications will also let us examine the different assumptions made by each (e.g., time heterogeneity, within/between variation). The next section will discuss the interpretations and results with this in mind.
\section{Results and Interpretation}
\subsection{Panel Regressions}
We begin by estimating the panel regression specifications given above. The results are shown below in Table \ref{table:panel}.
\begin{table}[H] \centering
    \caption{Panel regression results}
    \label{table:panel}
    \resizebox{\linewidth}{!} {
        \begin{tabular}{@{\extracolsep{5pt}}lcccc}
            \\[-1.8ex]\hline
            \hline \\[-1.8ex]
            PISA Math in global P99 & -31.228$^{*}$ & -38.928$^{*}$ & -34.569$^{*}$ & -115.492$^{***}$ \\
            & (16.314) & (20.802) & (17.850) & (36.599) \\
            IMO score per log population & 10.674$^{*}$ & 4.924$^{}$ & 3.122$^{}$ & -11.158$^{}$ \\
            & (6.363) & (7.470) & (6.441) & (15.728) \\
            ARWU insitutions & -166.641$^{**}$ & -162.705$^{*}$ & -176.049$^{**}$ & -159.960$^{}$ \\
            & (68.800) & (86.871) & (72.281) & (112.884) \\
            GDP per capita & & -0.001$^{}$ & -0.001$^{}$ & 0.003$^{}$ \\
            & & (0.001) & (0.001) & (0.003) \\
            PISA Math & & 0.619$^{}$ & 0.425$^{}$ & 0.178$^{}$ \\
            & & (0.716) & (0.684) & (1.735) \\
            Primary School Completion Rate & & -12.285$^{***}$ & -6.467$^{*}$ & -3.713$^{}$ \\
            & & (4.104) & (3.586) & (4.965) \\
            Lower Sec. Completion Rate & & -0.425$^{}$ & -0.953$^{}$ & -0.194$^{}$ \\
            & & (3.533) & (2.958) & (4.475) \\
            Upper Sec. Completion Rate & & 6.987$^{**}$ & 5.721$^{**}$ & 3.474$^{}$ \\
            & & (2.843) & (2.375) & (3.707) \\
            Democracy Rating & & 3.880$^{}$ & -5.350$^{}$ & 53.356$^{}$ \\
            & & (17.079) & (14.344) & (62.249) \\
            Population & & -0.000$^{}$ & -0.000$^{}$ & 0.000$^{}$ \\
            & & (0.000) & (0.000) & (0.000) \\
            Time Effects & No & No & Yes & Yes \\
            Fixed Effects & No & No & No & Yes \\
            Controls & No & Yes & Yes & Yes \\
            Entities & 89 & 89 & 89 & 89 \\
            \hline \\[-1.8ex]
            Observations & 440 & 440 & 440 & 440 \\
            $R^2$ & 0.037 & 0.077 & 0.374 & 0.553 \\
            Adjusted $R^2$ & 0.031 & 0.056 & 0.351 & 0.414 \\
            Residual Std. Error & 445.700 (df=436) & 439.888 (df=429) & 364.790 (df=423) & 346.511 (df=335) \\
            F Statistic & 5.616$^{***}$ (df=3; 436) & 3.589$^{***}$ (df=10; 429) & 15.813$^{***}$ (df=16; 423) & 3.983$^{***}$ (df=104; 335) \\
            \hline
            \hline \\[-1.8ex]
            \textit{Note:} & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\
            \end{tabular}
    }
\end{table}

@@

\subsection{Does IMO performance measure something else? (PCA)}
To explore whether IMO performance indeed captures a different set of unobservable factors (the cultural and institutional/structural elite orientation of an education system) compared to PISA math 99 share and ARWU institution count. Principle Components Analysis (PCA) may help analyze this by projecting the 3 variables onto 2 orthogonal vectors\footnote{Scikit-learn's implementation was used}. 2 components was found to be the MLE solution. See table \ref{table:pca} for the principle axes (parallel to eigenvectors).
\begin{table}[H]
    \caption{Principal axes in feature space}
    \label{table:pca}
    \centering
    \begin{tabular}{lrr}
        \toprule
         & Component A & Component B \\
        \midrule
        PISA math 99 & 0.126142 & 0.987521 \\
        ARWU ranked per MM pop. & 0.000565 & 0.094978 \\
        IMO score per log pop. & 0.992012 & -0.125625 \\
        \bottomrule
    \end{tabular}
\end{table}

These PCA results provide evidence that indeed IMO performance has captured a different set of unobservable factors (which I hypothesize include the elite orientation of an education system). The results indicate that when decompositing into 2 components, IMO score is in a separate component orthogonal to the one containing PISA math 99. ARWU ranked per MM pop. is not really a significant contributor to either.

@@

\begin{figure}[H]
    \caption{Shapley values for upper sec. completion rates}
    \centering
    \includegraphics[width=0.6\textwidth]{../shap_upper_sec.png}
    \label{fig:shap-upper-sec}
\end{figure}

The relationships between the controls and the elite orientation variables are likely to have changed as the XGBoost model has to some extent forced the existence of a relationship. For instance, it appears that IMO score is the most important feature in predicting upper secondary school completion rates using Shapley values (see figure \ref{fig:shap-upper-sec})\footnote{Shapley values can be used to help explain ML models}.
